True versus apparent shapes of astrophysical bow shocks

### 2016-06-15 papers/Paper1/sec-conic.tex
New figure with plane-of-sky projection

\section{Conic section approximation}
\label{sec:conic}

@@

In this section we will analyze the case where the resultant shape  of the bow shock is a conic curve (circle, ellipse, parabola or hyperbola).
These curves are mathematical simple to model and give us a good reference to understand the effects of the projection effects
described in the last section on other bow shocks. The source of the inner wind is located at the origin, and the center of the conic is located at
a distance $x_0$ from the source.

%Instead of the excentricity, we utilize the parameter $\theta_c$ to characterize the different curves, where
%$\tan\theta_c = \frac{b}{a}$,  $b$ and $a$ are the typical parameters of conics. A positive value for $\theta_c$ indicates that the given curve is a closed one, i.e
%an ellipse, while a negative value indicates that is an hyperbola. %Insert figures if neccesary  
\begin{figure}
\setlength\tabcolsep{0pt}
\begin{tabular}{cc}
\includegraphics[height=0.44\linewidth, trim=30 0 30 0, clip]{ellipse_py_lab} &
\includegraphics[height=0.44\linewidth, trim=60 0 70 0, clip]{hyperbola_py_lab}
\end{tabular}
\label{fig:conics}
\caption{Schematic representation of conic like bowshocks: A) Ellipse and B) Hyperbola The center of the conic is located at a distance $x_0=R_0-a$ from the origin at the $x$ axis. 
The radius of curvature at the axis is $R_c$, and $R_{90}$ is t he radius of the conic in the perpendicular direction to $R_0$.
}
\end{figure}
\begin{figure}
\includegraphics[width=\linewidth]{conic1}
\caption{A whole set of conics with the same radius of curvature. The parameter $\theta_c$ is used as the equivalent of the excentricity. This angle is defined as positive for closed curves
and negative for the opened ones. An special case is $\theta_c=0$, which describes a parabola.}
\label{fig:conics-family}
\end{figure}

Due to the similarities between the parametrization of the ellipse and the hyperbola, we can do the following parametrization:
\begin{align}
x = a\Cos(t)-x_0 \label{eq:par-x}\\ 
y = b\Sin(t) \label{eq:par-y}
\end{align}
where:
\begin{align}
\Cos(t) = \left\lbrace \begin{array}{c}
\cos t ~\mathrm{if~ellipse} \\
\cosh t ~\mathrm{if~hyperbola}
\end{array}\right.\\
\Sin(t) = \left\lbrace \begin{array}{c}
\sin t ~\mathrm{if~ellipse} \\
\sinh t ~\mathrm{if~hyperbola}
\end{array}\right. \\
-\pi < t < \pi \\
R_0 = a - x_0 \\
a<0 ~ \mathrm{for~hyperbola}
\end{align}
We have set a whole family of curves with a singular parameter denoted as $\tan\theta_c \equiv \frac{b}{a}$. Positive values of this parameter describe closed curves
(i.e. ellipses) and the negative ones describe open curves (i.e. hyperbolas). Particular cases are $\theta_c =\frac{\pi}{4}$, which describes a circle and $\theta_c=0$, which describes
a parabola.
To determine uniquely a given conic in terms of measurable quantities, we use the set of characteristic radii $(R_0,R_c,R_{90})$ in such way the conic parameters $(a,b,\theta_c)$
could be derived from the former. The transformation between the conic parameters and the characteristic radii are given by:
\begin{align}
\frac{R_c}{D} &= \frac{b^2}{a} \label{eq:Rc-conic}\\ 
\frac{R_{90}}{D} &= b\left(1 \mp \frac{(R_0-a)^2}{a^2}\right)
\end{align}
The inverse transformation is given by:
\begin{align}
a &= \frac{R_c}{2A-B^2} \label{eq:a-conic}\\
b &= \left(aR_c\right)^{1/2}\\
\theta_c &= \left|B^2 - 2A\right|^{1/2} \label{eq:thc-conic} \\
\mathrm{where:}& B\equiv \frac{R_{90}}{R_0} \\
A \equiv \frac{R_c}{R_0}
\end{align}
From equations (\ref{eq:a-conic}) we can see that for $R^2_{90} > 2 R_c$, the conic parameter $a$ is negative, so in this case the resulting conic is an hyperbola, and we must take the negative root
in equation (\ref{eq:thc-conic}). The opposite applies for ellipses.
\subsection{Plane-of-sky projection of conic sections} 
\begin{figure*}
\includegraphics[width=0.9\linewidth]{annotated}
\caption{Projected Radius of curvature vs projected $R_0$ for $\theta_c=60^\circ,45^\circ,28^\circ, 0^\circ,-15^\circ,-30^\circ$. Each colored points represent
the same shape at a different inclination. Being the dark blue the ones with the lower inclination and the red ones with the highest value. For closed shapes, there is
an assymptotic limit for the projected radius of curvature at high inclinations, while for open shapes there is an upper limit for inclination where a tangent line can be observed.}
\label{fig:Apqp}
\end{figure*}

>>>>

\section{Quadrics of revolution}
\label{sec:conic}

@@

For an arbitrary surface of revolution, application of
equations~(\ref{eq:tanphi}, \ref{eq:tangential}) to determine the
projected shape of the tangent line is not straightforward and in
general requires numerical techniques.  However, analytical results
can be found for the important class of surfaces known as
\textit{quadrics of revolution} \citep{Goldman:1983a, Gfrerrer:2009a},
which are formed by rotating a conic section plane curve about its
symmetry axis.  Examples are the sphere, spheroids (oblate and
prolate), and right circular paraboloids and hyperboloids.\footnote{We
  consider only the case of a single sheet of a 2-sheet hyperboloid or
  paraboloid.}  We ignore the degenerate cases of cylinders, cones,
and pairs of parallel planes.  While mathematically simple, these
quadrics are sufficiently flexible that they can provide a useful
approximation to more complex bow shock shapes.

The 
In this section we will analyze the case where the resultant shape  of the bow shock is a conic curve (circle, ellipse, parabola or hyperbola).
These curves are mathematical simple to model and give us a good reference to understand the effects of the projection effects
described in the last section on other bow shocks. The source of the inner wind is located at the origin, and the center of the conic is located at
a distance $x_0$ from the source.

%Instead of the excentricity, we utilize the parameter $\theta_c$ to characterize the different curves, where
%$\tan\theta_c = \frac{b}{a}$,  $b$ and $a$ are the typical parameters of conics. A positive value for $\theta_c$ indicates that the given curve is a closed one, i.e
%an ellipse, while a negative value indicates that is an hyperbola. %Insert figures if neccesary  
\begin{figure}
\setlength\tabcolsep{0pt}
\begin{tabular}{cc}
\includegraphics[height=0.54\linewidth, trim=30 0 30 0, clip]{ellipse_py_lab} &
\includegraphics[height=0.54\linewidth, trim=60 0 70 0, clip]{hyperbola_py_lab}
\end{tabular}
\label{fig:conics}
\caption{Schematic representation of conic like bowshocks: A) Ellipse and B) Hyperbola The center of the conic is located at a distance $x_0=R_0-a$ from the origin at the $x$ axis. 
The radius of curvature at the axis is $R_c$, and $R_{90}$ is t he radius of the conic in the perpendicular direction to $R_0$.
}
\end{figure}
\begin{figure}
\includegraphics[width=\linewidth]{conic1}
\caption{A whole set of conics with the same radius of curvature. The parameter $\theta_c$ is used as the equivalent of the excentricity. This angle is defined as positive for closed curves
and negative for the opened ones. An special case is $\theta_c=0$, which describes a parabola.}
\label{fig:conics-family}
\end{figure}

Due to the similarities between the parametrization of the ellipse and the hyperbola, we can do the following parametrization:
\begin{align}
x = a\Cos(t)-x_0 \label{eq:par-x}\\ 
y = b\Sin(t) \label{eq:par-y}
\end{align}
where:
\begin{align}
\Cos(t) = \left\lbrace \begin{array}{c}
\cos t ~\mathrm{if~ellipse} \\
\cosh t ~\mathrm{if~hyperbola}
\end{array}\right.\\
\Sin(t) = \left\lbrace \begin{array}{c}
\sin t ~\mathrm{if~ellipse} \\
\sinh t ~\mathrm{if~hyperbola}
\end{array}\right. \\
-\pi < t < \pi \\
R_0 = a - x_0 \\
a<0 ~ \mathrm{for~hyperbola}
\end{align}
We have set a whole family of curves with a singular parameter denoted as $\tan\theta_c \equiv \frac{b}{a}$. Positive values of this parameter describe closed curves
(i.e. ellipses) and the negative ones describe open curves (i.e. hyperbolas). Particular cases are $\theta_c =\frac{\pi}{4}$, which describes a circle and $\theta_c=0$, which describes
a parabola.
To determine uniquely a given conic in terms of measurable quantities, we use the set of characteristic radii $(R_0,R_c,R_{90})$ in such way the conic parameters $(a,b,\theta_c)$
could be derived from the former. The transformation between the conic parameters and the characteristic radii are given by:
\begin{align}
\frac{R_c}{D} &= \frac{b^2}{a} \label{eq:Rc-conic}\\ 
\frac{R_{90}}{D} &= b\left(1 \mp \frac{(R_0-a)^2}{a^2}\right)
\end{align}
The inverse transformation is given by:
\begin{align}
a &= \frac{R_c}{2A-B^2} \label{eq:a-conic}\\
b &= \left(aR_c\right)^{1/2}\\
\theta_c &= \left|B^2 - 2A\right|^{1/2} \label{eq:thc-conic} \\
\mathrm{where:}& B\equiv \frac{R_{90}}{R_0} \\
A \equiv \frac{R_c}{R_0}
\end{align}
From equations (\ref{eq:a-conic}) we can see that for $R^2_{90} > 2 R_c$, the conic parameter $a$ is negative, so in this case the resulting conic is an hyperbola, and we must take the negative root
in equation (\ref{eq:thc-conic}). The opposite applies for ellipses.
\subsection{Plane-of-sky projection of quadric surfaces} 
\begin{figure*}
\includegraphics[width=0.9\linewidth]{annotated}
\caption{Projected Radius of curvature vs projected $R_0$ for $\theta_c=60^\circ,45^\circ,28^\circ, 0^\circ,-15^\circ,-30^\circ$. Each colored points represent
the same shape at a different inclination. Being the dark blue the ones with the lower inclination and the red ones with the highest value. For closed shapes, there is
an assymptotic limit for the projected radius of curvature at high inclinations, while for open shapes there is an upper limit for inclination where a tangent line can be observed.}
\label{fig:Apqp}
\end{figure*}

### 2016-06-15 papers/Paper1/sec-projection.tex
New figure with plane-of-sky projection

In this section we calculate the apparent shape of a limb brightened  border of a generic bow shock with cilindrical geometry, observed in the
plane of the sky.

@@

We can specify the shell's position in cartesian coordinates $(x,y,z)$, where the x axis points towards the symetry axis of the system.
Since the shell is cylindrically symmetric, we can describe it's shape fully with the function $R(\theta)$, such that:
\begin{equation}
\left(\begin{array}{c}
x \\ y \\ z
\end{array}
\right) = R(\theta)\left(\begin{array}{c}
\cos\theta \\
\sin\theta\cos\phi \\
\sin\theta\sin\phi
\end{array}\right)
\end{equation} 
where $\theta$ is the polar angle and $\phi$ the azimutal angle from spherical coordinates

The observer's reference frame is denotated with primes, and is rotated respect y axis by an angle $i$. The transformation between the shell frame and the observed
frame is given by:
\begin{equation}
\left(\begin{array}{c}
x' \\ y' \\ z'
\end{array}
\right) = \left(\begin{array}{c}
x\cos i - z\sin i\\
y \\
z\cos i + x\sin i
\end{array}\right)
\label{eq:Trans}
\end{equation}

@@

The normal and tangent vectors to the shell's border in the shell's frame are given by:

\begin{align}
\hat{t} = \left(\begin{array}{c}
-\cos\alpha \\
\sin\alpha\cos\phi\\
\sin\alpha\sin\phi
\end{array}\right)\\
\hat{n} = \left(\begin{array}{c}
\sin\alpha \\
\cos\alpha\cos\phi \\
\cos\alpha\sin\phi
\end{array}\right)
\end{align}

Where:

\begin{align}
\tan\alpha = -\left.\frac{dy}{dx}\right|_{R(\theta)} \\
\label{eq:tanalpha}
\end{align}

Then we apply the transformation (\ref{eq:Trans}) to the normal and tangent vectors to obtain:

\begin{align}
\hat{n}' &= \left(\begin{array}{c}
(\cos\theta+\omega\sin\theta)\cos i -(\sin\theta-\omega\cos\theta)\sin i \sin\phi\\
(\sin\theta-\omega\cos\theta)\cos\phi \\
(\cos\theta+\omega\sin\theta)\sin i + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
\end{array}\right)\\
\hat{t}' &= \left(\begin{array}{c}
-(\sin\theta-\omega\cos\theta)\cos i - (\cos\theta+\omega\sin\theta)\sin\phi\sin i \\
(\cos\theta+\omega\sin\theta)\cos\phi \\
-(\cos\theta+\omega\sin\theta)\sin i + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
\end{array}\right)
\end{align}

Where $\omega = \frac{1}{R}\frac{dR}{d\theta}$

In the thin shell case, the limb brightened border of the shell is such that $\hat{n}'\cdot \hat{z}'$. 
The values for $\phi$ that satisfy this relation for each inclination $i$ are given by:

\begin{equation}
\sin\phi_t = \tan i\tan\alpha = \tan i \frac{1+\omega\tan\theta}{\omega-\tan\theta}
\label{eq:tanphi}
\end{equation}

With this, the coordinates of the limb brightened shell are given by:

\begin{equation}
\left(\begin{array}{c}
x'_t \\ y'_t \\ z'_t
\end{array}\right)= R(\theta)\left(\begin{array}{c}
\cos\theta\cos i - \sin\theta\sin\phi_t \sin i \\
\sin\theta(1-\sin^2\phi_t)^{1/2} \\
\cos\theta\sin i +\sin\theta\sin\phi_t\cos i
\end{array}\right)
\label{eq:tangential}
\end{equation}

It is important to note that the equation (\ref{eq:tanphi}) does not have a solution for arbitrary values for $\theta$ and $i$, since
it's required that $|\sin\phi_t|<1$. If $i\neq 0$, then, the  allowed values for $\theta$ are such that $\theta > \theta_\parallel$, where
$\theta_\parallel$ is given implicitly by:

\begin{align}
\tan\theta_\parallel = \frac{|\tan i| + \omega(\theta_\parallel)}{1-\omega(\theta_\parallel) |\tan i|}
\label{eq:thetapar}
\end{align}

@@

Considering further applications to bow shocks, we will consider open shells. In order to compare the shell shape given by $R(\theta)$ with observations,
it is convenient to define the following apparent radii in the observer frame: $R'_\parallel$ and $R'_\perp$. These are projected distances of the shell tangent line
from the origin. The first is measured in the direction of the symetry axis, and the second in a perpendicular direction. More concretely $R'_\parallel = x'_t(y'_t=0)$
and $R'_\perp = y'_t(x'_t=0)$. From equations (\ref{eq:tanphi}) and (\ref{eq:tangential}) we find that:

\begin{align}
R'_\parallel = R(\theta_\parallel)\cos(\theta + i) \label{eq:Rpar} 
\end{align}

Where $\theta_\parallel$ is the solution of equation (\ref{eq:thetapar}), and

\begin{align}
R'_\perp = R(\theta_\perp)\sin\theta_\perp\left(1-\sin^2(\phi_t(\theta_\perp))\right)^{1/2}
\end{align}

Where $\theta_\perp$ is the solution of the next implicit equation:

\begin{align}
\cot\theta_\perp = \frac{1-\left(1+\omega(\theta_\perp)^2\sin^22i\right)^{1/2}}{2\omega(\theta_\perp\cos^2 i)}
\end{align}

>>>>

In this section we calculate the apparent shape on the plane of the
sky of the limb brightened border of a shock or shell that is
idealized as as arbitrary cylindrically symmetric surface.

@@

Consider body-frame cartesian coordinates $(x,y,z)$, where \(x\) is
the symmetry axis, and spherical polar coordinates
\((R, \theta, \phi)\), where \(\theta\) is the polar angle and
\(\phi\) the azimuthal angle.  Since the surface is cylindrically
symmetric, it is can be specified as $R = R(\theta)$, so that
cartesian coordinates on the surface are:
\begin{equation}
\left(\begin{array}{c}
x \\ y \\ z
\end{array}
\right) = R(\theta)\left(\begin{array}{c}
\cos\theta \\
\sin\theta\cos\phi \\
\sin\theta\sin\phi
\end{array}\right).
\end{equation} 
Suppose that the viewing direction makes an angle \(i\) with the \(z\)
axis, so that we can define observer-frame coordinates
\((x', y', z')\), which are given by the rotating the body-frame
coordinates:
\begin{equation}
\left(\begin{array}{c}
x' \\ y' \\ z'
\end{array}
\right) = \left(\begin{array}{c}
x\cos i - z\sin i\\
y \\
z\cos i + x\sin i
\end{array}\right).
\label{eq:Trans}
\end{equation} 
In what follows, all quantities in the observer's frame are denoted by
attaching a prime to the equivalent quantity in the body frame.  The
coordinates \(x'\) and \(y'\) are in the plane of the sky, with \(x'\)
being the projected symmetry axis of the surface, while the line of
sight lies along \(-z'\).  The inclination angle \(i\) is defined so
that \(i = 0^\circ\) when the surface is viewed perpendicular to its
axis (\textit{side on}) and \(i = 90^\circ\) when it is viewed along
its axis (\textit{end on}).

@@

The normal and tangent vectors to the shell's border in the shell's frame are given by:
\begin{align}
\hat{t} = \left(\begin{array}{c}
-\cos\alpha \\
\sin\alpha\cos\phi\\
\sin\alpha\sin\phi
\end{array}\right)\\
\hat{n} = \left(\begin{array}{c}
\sin\alpha \\
\cos\alpha\cos\phi \\
\cos\alpha\sin\phi
\end{array}\right)
\end{align}
Where:
\begin{align}
\tan\alpha = -\left.\frac{dy}{dx}\right|_{R(\theta)} \\
\label{eq:tanalpha}
\end{align}
Then we apply the transformation (\ref{eq:Trans}) to the normal and tangent vectors to obtain:
\begin{align}
\hat{n}' &= \left(\begin{array}{c}
(\cos\theta+\omega\sin\theta)\cos i -(\sin\theta-\omega\cos\theta)\sin i \sin\phi\\
(\sin\theta-\omega\cos\theta)\cos\phi \\
(\cos\theta+\omega\sin\theta)\sin i + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
\end{array}\right)\\
\hat{t}' &= \left(\begin{array}{c}
-(\sin\theta-\omega\cos\theta)\cos i - (\cos\theta+\omega\sin\theta)\sin\phi\sin i \\
(\cos\theta+\omega\sin\theta)\cos\phi \\
-(\cos\theta+\omega\sin\theta)\sin i + (\sin\theta-\omega\cos\theta)\sin\phi\cos i
\end{array}\right)
\end{align}
Where $\omega = \frac{1}{R}\frac{dR}{d\theta}$

In the thin shell case, the limb brightened border of the shell is such that $\hat{n}'\cdot \hat{z}'$. 
The values for $\phi$ that satisfy this relation for each inclination $i$ are given by:
\begin{equation}
\sin\phi_t = \tan i\tan\alpha = \tan i \frac{1+\omega\tan\theta}{\omega-\tan\theta}
\label{eq:tanphi}
\end{equation}
With this, the coordinates of the limb brightened shell are given by:
\begin{equation}
\left(\begin{array}{c}
x'_t \\ y'_t \\ z'_t
\end{array}\right)= R(\theta)\left(\begin{array}{c}
\cos\theta\cos i - \sin\theta\sin\phi_t \sin i \\
\sin\theta(1-\sin^2\phi_t)^{1/2} \\
\cos\theta\sin i +\sin\theta\sin\phi_t\cos i
\end{array}\right)
\label{eq:tangential}
\end{equation}

It is important to note that the equation (\ref{eq:tanphi}) does not have a solution for arbitrary values for $\theta$ and $i$, since
it's required that $|\sin\phi_t|<1$. If $i\neq 0$, then, the  allowed values for $\theta$ are such that $\theta > \theta_\parallel$, where
$\theta_\parallel$ is given implicitly by:
\begin{align}
\tan\theta_\parallel = \frac{|\tan i| + \omega(\theta_\parallel)}{1-\omega(\theta_\parallel) |\tan i|}
\label{eq:thetapar}
\end{align}

@@

Considering further applications to bow shocks, we will consider open shells. In order to compare the shell shape given by $R(\theta)$ with observations,
it is convenient to define the following apparent radii in the observer frame: $R'_\parallel$ and $R'_\perp$. These are projected distances of the shell tangent line
from the origin. The first is measured in the direction of the symetry axis, and the second in a perpendicular direction. More concretely $R'_\parallel = x'_t(y'_t=0)$
and $R'_\perp = y'_t(x'_t=0)$. From equations (\ref{eq:tanphi}) and (\ref{eq:tangential}) we find that:
\begin{align}
R'_\parallel = R(\theta_\parallel)\cos(\theta + i) \label{eq:Rpar} 
\end{align}
Where $\theta_\parallel$ is the solution of equation (\ref{eq:thetapar}), and
\begin{align}
R'_\perp = R(\theta_\perp)\sin\theta_\perp\left(1-\sin^2(\phi_t(\theta_\perp))\right)^{1/2}
\end{align}
Where $\theta_\perp$ is the solution of the next implicit equation:
\begin{align}
\cot\theta_\perp = \frac{1-\left(1+\omega(\theta_\perp)^2\sin^22i\right)^{1/2}}{2\omega(\theta_\perp\cos^2 i)}
\end{align}
